---
title: "Auto Production in the United States: Time Series Analysis"
author: "Paige Gonzales"
date: '2023-07-28'
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r message=FALSE, warning=FALSE, include=FALSE}

#Load necessary libraries
library(dplyr)
library(forecast)
library(astsa)
library(TSA)
library(zoo)
library(ggplot2)
library(scales)

#Suppress warnings and messages
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

```

# 1. Original forecast as of December 2022

## a. Read in data

```{r}

#Load the data file
autoProd <- read.csv("DAUPNSAOriginal.csv")

#Verify header and dimensions
head(autoProd)
dim(autoProd)

#Check for any missing values
sum(is.na(autoProd))

```

## b. Convert to time series data

```{r}

#Change the data to date format
autoProd$DATE <- as.Date(autoProd$DATE)

#Convert to time series
autoProdTS <- ts(data=autoProd$DAUPNSA,frequency=12,start=c(1993,1),end=c(2022,9))

#Plot time series for visualizing
ts.plot(autoProdTS, ylab="Domestic Auto Production-Thousand Units")

```

## c. Check for stationarity

```{r}

#Plot ACF of original data
#Notice not stationary
#Time series plot shows a trend and non-constant variance
acf2(autoProdTS)

#Stabilize variance using log 
logCars <- log(autoProdTS)

#Re-check time series plot
ts.plot(logCars, ylab="Domestic Auto Production-Thousand Units")

#Re-check ACF plot
acf2(logCars)

#Remove trend by first order differencing
detrendCars <- diff(logCars)

#Check time series plot again
ts.plot(detrendCars, ylab="Domestic Auto Production-Thousand Units")

#Check ACF/PACF again
#Non-stationary due to seasonality
acf2(detrendCars)  

#Seasonality check - shows yearly seasonality
ggseasonplot(autoProdTS, main = "Seasonal Plot of Domestic Auto Production", ylab = "Production (Thousands)")

#Seasonal differencing added to ACF plot
acf2(diff(detrendCars,12), 60)

```

## d. Seasonal modeling (SARIMA) and forecast

```{r}

#SARIMA model and diagnostics
sarima((logCars),0,1,2,0,1,1,12)

#Suggested model by autoarima() and diagnostics
auto.arima(logCars)
sarima(logCars,1,1,1,0,0,2,12)

#Forecast for next 12 months using user created SARIMA model 
forecastOriginal <- sarima.for(autoProdTS,12,0,1,2,0,1,1,12)

#Predictions for next 12 months (starting (October 2022))
predOriginal <- forecastOriginal$pred

```

# 2. Compare originally predicted values to actual values from October 2022 to May 2023

## a. Read in data

```{r}

#Load the data file
autoProdUpdate <- read.csv("DAUPNSAUpdated.csv")

#Verify header and dimensions
head(autoProdUpdate)
dim(autoProdUpdate)

#Check for any missing values
sum(is.na(autoProdUpdate))

```

## b. Convert new data to time series

```{r}

#Convert to date
autoProdUpdate$DATE <- as.Date(autoProdUpdate$DATE)

#Change to time series data
autoProdUpdateTS <- ts(data=autoProdUpdate$DAUPNSA,frequency=12,start=c(1993,1),end=c(2023,5))

```

## c. Compare new data to originally predicted values

```{r}

#Select only values predicted originally that have values that are currently available (October 2022 to May 2022)
actualValues <- window(autoProdUpdateTS, start = 2022 + 9/12, end = 2023 + 4/12)
predOriginalValues <- window(predOriginal, start = 2022 + 9/12, end = 2023 + 4/12)

#Calculate percent difference between predicted and actual values
percentDiff <- ((predOriginalValues - actualValues)/actualValues)*100

# Data frame for plotting
data <- data.frame(
  time = as.Date(time(actualValues)),
  predicted = predOriginalValues,
  actual = actualValues,
  percent_diff = percentDiff
)

#Manually set colors for plot
colors <- c("Actual" = "blue", "Predicted" = "red")

#Plot predicted to actual for original project
ggplot(data, aes(x = time)) +
  geom_line(aes(y = predicted, color = "Predicted"), linetype = 2) +
  geom_line(aes(y = actual, color = "Actual")) +
  geom_point(aes(y = predicted, color = "Predicted"), size = 2) +
  geom_point(aes(y = actual, color = "Actual"), size = 2) +
  geom_text(aes(y = predicted, label = paste0(round(percent_diff, 2), "%")), nudge_y = 12, nudge_x = 5, color = "red", size = 3, check_overlap = TRUE) +
  labs(title = "U.S. Domestic Auto Prediction - Predicted vs. Actual", x = "Month/Day", y = "Value", color = "Legend") +
  theme_classic() + 
  theme(plot.margin = margin(40, 30, 30, 30), axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_date(labels = date_format("%b %y"), breaks = "1 month") + 
  scale_color_manual(values = colors)

```

# 3. Adjustment for COVID-19 outliers 

## a. Remove outliers

```{r}

#Make a copy of the original dataset 
autoProdCOVIDTS <- autoProdTS

#Replace two major outliers due to COVID with NA
autoProdCOVIDTS[328] <- NA
autoProdCOVIDTS[329] <- NA

#Smooth missing values - library(zoo)
autoProdCOVIDTS <- na.approx(autoProdCOVIDTS)

#Windows starting in 2019 for plotting adjusted and actual values
autoProdCOVIDTSWindow <- window(autoProdCOVIDTS, start = 2019 + 1/12, end = 2023 + 0/12)
autoProdTSWindow <- window(autoProdTS, start = 2019 + 1/12, end = 2023 + 0/12)

#Date frame for plotting time series comparing adjusted to actual values
dataCovidTS <- data.frame(
  time = as.Date(time(autoProdCOVIDTSWindow)),
  adjusted = autoProdCOVIDTSWindow,
  actual = autoProdTSWindow
)

#Manually set colors for plot
colors <- c("Actual" = "blue", "Adjusted" = "red")

#Plot comparing adjusted values to actual values
ggplot(dataCovidTS, aes(x = time)) +
  geom_line(aes(y = adjusted, color = "Adjusted"), linetype = 2) +
  geom_line(aes(y = actual, color = "Actual")) +
  geom_point(aes(y = adjusted, color = "Adjusted"), size = 1) +
  geom_point(aes(y = actual, color = "Actual"), size = 1) +
  labs(title = "Real Values Versus Smoothed Values", x = "Month/Day", y = "Value", color = "Legend") +
  theme_classic() + 
  theme(plot.margin = margin(40, 30, 30, 30), axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_date(labels = date_format("%b %y"), breaks = "3 month") + 
  scale_color_manual(values = colors)

```

## b. Re-run data with the original model

```{r}

#ACF of original data
acf2(autoProdCOVIDTS)

#Stabilize variance 
logCarsCOVID <- log(autoProdCOVIDTS)

#Check ACF after stabilizing variance
#Needs to be detrended
acf2(logCarsCOVID)

#Remove trend
adjustCarsCOVID <- diff(logCarsCOVID)

#Check ACF/PACF
#Non-stationary due to seasonality
acf2(adjustCarsCOVID)

#Seasonality check
ggseasonplot(autoProdCOVIDTS)

#Seasonal differencing for annual
acf2(diff(adjustCarsCOVID,12), 60)

#SARIMA model and diagnostics (original model)
sarima(logCarsCOVID,0,1,2,0,1,1,12)

#Prediction for the next 12 months using original model but with COVID outliers removal
forecastCOVID <- sarima.for(autoProdCOVIDTS,12,0,1,2,0,1,1,12)

```

# 4. Compare Results Of Original Predictions to Actual and Smoothed

```{r}

#Select predictions only 
predCOVID <- forecastCOVID$pred

#Select only the prediction that there are actual values for (October 2022 - May 2023)
predCOVIDValues <- window(predCOVID, start = 2022 + 09/12, end = 2023 + 4/12)

dataCOVID <- data.frame(
  time = as.Date(time(actualValues)),
  predicted = predOriginalValues,
  actual = actualValues,
  covid = predCOVIDValues
)

#Manually set colors for plot
colors <- c("Actual" = "blue", "Pre-Smoothing" = "red", "Post-Smoothing" = "green")

#Plot
ggplot(dataCOVID, aes(x = time)) +
  geom_line(aes(y = predicted, color = "Pre-Smoothing")) +
  geom_line(aes(y = actual, color = "Actual")) +
  geom_line(aes(y = covid, color = "Post-Smoothing")) +
  geom_point(aes(y = predicted, color = "Pre-Smoothing"), size = 2) +
  geom_point(aes(y = actual, color = "Actual"), size = 2) +
  geom_point(aes(y = covid, color = "Post-Smoothing"), size = 2) +
  labs(title = "A Comparison of Actual Values to Pre-Smoothing and Post-Smoothing Predictions", x = "Month/Day", y = "Auto Production (thousands)", color = "Legend") +
  theme_classic() + 
  theme(plot.margin = margin(40, 30, 30, 30), axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_date(labels = date_format("%b %y"), breaks = "1 month") + 
  scale_color_manual(values = colors)
```

# 5. Adjust model for COVID smoothed data

```{r}
#ACF of COVID smoothed data after 
acf2(diff(adjustCarsCOVID,12), 60)

#SARIMA model and diagnostics
sarima(logCarsCOVID,0,1,0,4,1,0,12)

#Suggested model by autoarima() and diagnostics
auto.arima(logCarsCOVID)
sarima(logCarsCOVID,3,1,1,0,0,2,12)

#None show good fit
```

# 6. Re-run original model with latest 8 months' values 

```{r}

#ACF of original data
acf2(autoProdUpdateTS)

#Stabilize variance 
logCarsUpdate <- log(autoProdUpdateTS)
acf2(logCarsUpdate)

#Remove trend
adjustCarsUpdate <- diff(logCarsUpdate)

#Check ACF/PACF
#Non-stationary due to seasonality
acf2(adjustCarsUpdate)

#Seasonality check
ggseasonplot(autoProdUpdateTS)

#Seasonal differencing
acf2(diff(adjustCarsUpdate,12), 60)

#SARIMA model and diagnostics
sarima(logCarsUpdate,0,1,2,0,1,1,12)

#Prediction for the next 12 months using original model 
sarima.for(autoProdUpdateTS, 12,0,1,2,0,1,1,12)
forecastUpdated <- sarima.for(autoProdUpdateTS, 12,0,1,2,0,1,1,12)

```




